---
output:
  word_document: default
  html_document: default
---
# Phase 1: Descriptive Analytics
## Dallas Nifong

```{r}
# Load Packages
library(tidyverse)
library(GGally)
library(ggcorrplot)
library(naniar)
library(skimr)
library(gridExtra)
```

```{r}
# Phase 1 chart themes
library(scales)

theme_phase1 <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey35", size = 11, margin = margin(b = 12)),
    plot.caption  = element_text(color = "grey50", size = 9, hjust = 0),
    axis.title    = element_text(color = "grey30"),
    panel.grid.minor = element_blank(),
    legend.position  = "top",
    plot.title.position = "plot"
  )

fail_cols <- c("No" = "#B0B7BC", "Yes" = "#C0392B")

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150,
                      out.width = "100%")
```


```{r}
# Load data
train <- read_csv("train.csv")
```

```{r}
# Convert categoricals to factors
train <- train %>%
  mutate(failure      = as_factor(failure),
         product_code = as_factor(product_code),
         attribute_0  = as_factor(attribute_0),
         attribute_1  = as_factor(attribute_1),
         attribute_2  = as_factor(attribute_2),
         attribute_3  = as_factor(attribute_3))
```

```{r}
# Structure
str(train)
```

```{r}
# Remove ID
train <- train %>% select(-id)
```


```{r}
# Summary
summary(train)
```

```{r}
# Bundle missingness and distribution shape
skim(train)
```

```{r}
# Target distribution
train %>% count(failure) %>% mutate(pct = n / sum(n))
```

```{r}
# Missingness Analysis

# Missingness bar

# Labels for display only, rename columns on a throwaway copy
train_labeled <- train
names(train_labeled) <- names(train_labeled) %>%
  str_replace_all("_", " ") %>%        # underscores -> spaces
  str_to_title()                       # Title Case

gg_miss_var(train_labeled) +
  labs(
    title    = "Missing data climbs with measurement number",
    subtitle = "Higher-numbered lab measurements have the most gaps",
    x = NULL,
    y = "Number of Missing Values"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_phase1 +
  theme(panel.grid.major.y = element_blank())

# Clustered map
set.seed(123)
train %>%
  slice_sample(n = 3000) %>%
  vis_miss(cluster = TRUE) +
  labs(title = "Missingness is scattered, not concentrated in certain products") +
  theme(axis.text.x = element_text(angle = 60, hjust = 0, size = 7))

# Overall & per-variable missing rates table
train %>%
  summarize(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(train), 2)) %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

# Rows with at least one missing value
train %>%
  summarize(rows_with_any_na = mean(!complete.cases(.))) %>%
  mutate(rows_with_any_na = round(100 * rows_with_any_na, 1))

# Missingness to failure
train %>%
  mutate(loading_missing = is.na(loading)) %>%
  group_by(loading_missing) %>%
  summarize(failure_rate = round(100 * mean(failure == "Yes"), 2),
            n = n())
```

```{r}
# Resolve missingness
library(mice)

# Loading is missing in <1% of rows, drop those rows.
train <- train %>% drop_na(loading)

# Remaining gaps (measurements, up to ~8-9%) filled with PMM via mice.
imp   <- mice(train, method = "pmm", m = 5, seed = 123, printFlag = FALSE)
train <- complete(imp)

# Confirm no missing values remain.
sum(is.na(train))
```


```{r}
# Loading split by failure
overall_med <- median(train$loading, na.rm = TRUE)

overall_med <- median(train$loading, na.rm = TRUE)

ggplot(train, aes(x = failure, y = loading, fill = failure)) +
  geom_hline(yintercept = overall_med, linetype = "dashed",
             color = "grey60", linewidth = 0.4) +
  geom_boxplot(width = 0.55, outlier.alpha = 0.15,
               outlier.size = 0.7, linewidth = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "grey20") +
  scale_fill_manual(values = fail_cols) +
  scale_x_discrete(labels = c("No" = "Passed", "Yes" = "Failed")) +
  labs(
    title    = "Failed products absorbed more fluid",
    subtitle = "Loading is the single strongest signal of failure (median 132 vs 120)",
    x = "Product outcome",
    y = "Loading (fluid absorbed)",
    caption  = "Diamond = mean.  Dashed line = overall median.  Outliers shown at low opacity."
  ) +
  theme_phase1 +
  theme(legend.position = "none")

train %>% group_by(failure) %>%
  summarize(mean_loading = mean(loading, na.rm = TRUE),
            median_loading = median(loading, na.rm = TRUE))
```

```{r}
quant_levels <- c("Loading",
                  paste("Measurement", 0:17))   # Loading, Measurement 0 ... 17 in order

train %>%
  select(failure, loading, starts_with("measurement_")) %>%
  pivot_longer(-failure, names_to = "variable", values_to = "value") %>%
  mutate(variable = str_to_title(str_replace_all(variable, "_", " ")),
         variable = factor(variable, levels = quant_levels)) %>%   # lock the order
  ggplot(aes(x = failure, y = value, fill = failure)) +
  geom_boxplot(outlier.alpha = 0.1, outlier.size = 0.5, linewidth = 0.4) +
  facet_wrap(~ variable, scales = "free_y", ncol = 5) +
  scale_fill_manual(values = fail_cols) +
  scale_x_discrete(labels = c("No" = "Passed", "Yes" = "Failed")) +
  labs(
    title    = "Loading is the only predictor that separates failures",
    subtitle = "Across all 19 quantitative measures, only loading shifts between passed and failed products",
    x = NULL, y = NULL
  ) +
  theme_phase1 +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 9))
```

```{r}
train %>%
  select(failure, product_code, attribute_0, attribute_1, attribute_2, attribute_3) %>%
  pivot_longer(-failure, names_to = "variable", values_to = "level") %>%
  mutate(
    variable = str_to_title(str_replace_all(variable, "_", " ")),
    level    = str_to_title(str_replace_all(level, "_", " "))   # "material_5" -> "Material 5"
  ) %>%
  ggplot(aes(x = level, fill = failure)) +
  geom_bar(position = "fill") +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  scale_fill_manual(values = fail_cols, labels = c("No" = "Passed", "Yes" = "Failed")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "None of the categorical predictors separate failures",
    subtitle = "Failure rate holds near 21% across every product code and attribute level",
    x = NULL, y = "Proportion", fill = NULL
  ) +
  theme_phase1 +
  theme(
    strip.text  = element_text(size = 9),
    axis.text.x = element_text(size = 8)
  )
```


```{r}
# Product code chart
pc_summary <- train %>%
  group_by(product_code) %>%
  summarize(fail_rate = mean(failure == "Yes"))

ggplot(pc_summary, aes(x = product_code, y = fail_rate)) +
  geom_col(fill = "#C0392B", width = 0.65) +
  geom_text(aes(label = scales::percent(fail_rate, accuracy = 0.1)),
            vjust = -0.5, size = 4, color = "grey20") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.28),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "Failure rate is nearly identical across product types",
    subtitle = "Which prototype a product is tells you little about whether it fails",
    x = "Product Code", y = "Failure Rate"
  ) +
  theme_phase1 +
  theme(panel.grid.major.x = element_blank())
```

## What the data tells us

**The amount of fluid a product absorbs is the clearest warning sign of failure.**
Products that absorbed more fluid are more likely to fail. On average, failed
products absorbed 132 units versus 120 for a product that passed, so a noticeable
difference.

**The individual lab tests, on their own, tell us little.** 
Of the eighteen lab measurements, none show a meaningful relationship with failure 
individually. This doesn't mean they are useless, combinations of them may still matter
when I build models in Phase 2.

**The type of product doesn't change the odds of failure.** 
This data covered five different product recipes, all five fail at roughly the
same rate (about 20-23%).

**The data is incomplete, so it was cleaned.**
More than half of all products were missing at least one lab measurement.
The small number of rows missing a "loading" value (under 1%) were dropped,
and the remaining gaps were filled using predictive mean matching.

## Bottom line

If Keep It Dry wants an early warning that a product might fail, I believe "loading" 
is the most promising signal found so far. The lab tests and the product type alone
don't say much, but they may matter in Phase 2 once a model can weigh several of them
together.